Comparing corrosion control treatments using a robust Bayesian generalized additive model

### 2022-05-24 README.Rmd
updated ref list

This repository contains the R code and other materials for the paper "Comparing corrosion control treatments using a robust Bayesian generalized additive model", currently available as a [preprint](http://doi.org/10.26434/chemrxiv-2022-b3f72). Bespoke functions used to fit the models are available as a separate package, [`bgamcar1`](https://github.com/bentrueman/bgamcar1).

>>>>

This repository contains the R code and other materials for the paper "Comparing corrosion control treatments using a robust Bayesian generalized additive model", currently available as a [preprint](http://doi.org/10.26434/chemrxiv-2022-b3f72-v2). Bespoke functions used to fit the models are available as a separate package, [`bgamcar1`](https://github.com/bentrueman/bgamcar1).

### 2022-05-24 Rmarkdown/paper.Rmd
updated ref list

Pipe loop studies are used routinely to evaluate corrosion control treatment, and updated regulatory guidance will ensure that they remain an important tool for water quality management. But they are inherently complex and the data they generate are difficult to analyze: non-linear time-trends, non-detects, extreme values, and autocorrelation are common features, the latter due to repeatedly measuring the effluent from test pipes. Popular statistical tests, such as the Student *t* or rank-sum tests, are often inadequate descriptions of the data. Here, we propose an approach to statistical analysis of pipe loop data that accommodates many of these difficult-to-model characteristics: a robust Bayesian generalized additive model with continuous-time autoregressive errors. Our model facilitates corrosion control treatment comparisons without the need for imputing non-detects or special handling of outliers. It is well-suited to describing nonlinear trends without overfitting, and it accounts for reduced information content due to autocorrelation. We demonstrate the model using an example pipe loop study comprising four years of data, multiple pipe configurations, and multiple orthophosphate dosing protocols. We compare the experimental treatments, finding that an initially high dose of orthophosphate (2 mg P L^-1^) that is subsequently lowered (0.75 mg P L^-1^) can yield lower lead release than an intermediate dose (1 mg P L^-1^) in the long term. We also find that---consistent with previous work---galvanic corrosion yields relatively high particulate lead release, especially at higher orthophosphate doses. An advantage of the Bayesian approach we adopt here is that it yields the full posterior distribution of all parameters and all quantities derived from those parameters, including the model predictions. This means that analysts can construct any comparison that is relevant to the study goals, without the need to rely on a particular statistical test.

@@

Pipe loop studies are used to model drinking water distribution systems, mainly for evaluating corrosion control treatment. The US EPA's Revised Lead and Copper Rule is expected to result in many water utilities conducting pipe loop studies [@estessmargiassi_understanding_2020], and the data they generate will need to be analyzed. But pipe loop study data are inherently complex and there is a real risk of misinterpreting them. The following are just some of the challenges that arise (Figure 1):

@@

All of the above present barriers to reliable statistical inference, but left-censoring and autocorrelation can be particularly challenging. Left-censoring is often addressed by substituting a constant for values below the detection limit and sometimes by removing the censored values. Neither approach is advisable because both introduce artificial patterns that can distort results, sometimes seriously [@helsel_statistics_2012]. Autocorrelation overstates the precision of statistical estimates [@gelman_bayesian_2014]. As a consequence, it inflates the probability of falsely rejecting null hypotheses [@helsel1992statistical]. The effect of autocorrelation can be surprisingly large: 100 observations from a time series may provide no greater precision than 10 independent observations if the 100 are strongly correlated in time [@helsel1992statistical]. That is, *n* autocorrelated observations provide less information than *n* independent observations, and the difference can be large if autocorrelation is strong [@gelman_bayesian_2014]. Accounting for autocorrelation can improve predictive accuracy, but if it is ignored, analysts run the risk of making incorrect decisions based on the data. And while rank-based procedures such as the rank-sum test can accommodate left-censoring, most are not designed for autocorrelated data [@helsel1992statistical; @helsel_statistics_2012].

@@

1. A global---possibly nonlinear---trend describing variation in the parameter of interest over time across all pipes.
2. A set of group-level trends that describe the deviations from the global trend particular to the individual pipes.
3. A global seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature. (This component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it.)
4. An autocorrelated error term.

We demonstrate this approach using an example pipe loop dataset representing four years of data, multiple pipe configurations, and multiple orthophosphate dosing protocols. Our model estimates autocorrelation, despite an irregular sampling frequency; it also accommodates censored observations without imputation and is robust against extreme analyte concentrations due to particle release---meaning that no special treatment of outliers is necessary. An advantage of our Bayesian approach is the ability to move beyond point estimates and intervals: Bayesian models yield the full posterior distribution of all parameters, and all quantities derived from those parameters, including the predicted values. This means that we can construct any comparison we wish to investigate, without the need to decide on a particular statistical test. Moreover, the use of additive smoothing splines permits a description of time series using multiple nonlinear components, which can yield a more nuanced understanding of differences in lead concentrations over time.

@@

Lead, iron, aluminum, and manganese were measured by ICP-MS [@sm_3125] with method detection limits estimated at 0.2, 5, 5, and 2 µg L^-1^, respectively. An aliquot of each sample was filtered at 0.45 µm using a teflon filter and a vacuum-filter apparatus. Temperature was measured using an infrared thermometer, total chlorine [@sm_4500_cl_g] and orthophosphate [@sm_4500_p_e] were measured using a spectrophotometer, and pH was measured using a benchtop meter. Water quality---including alkalinity [@sm_2320b], true colour [@sm_2120c], conductivity [@sm_2510b], fluoride [@sm_4500], total hardness [@sm_2340], and turbidity [@sm_2130b] ---is summarized in Table S4. Colour was quantified after filtration at 0.45 µm, using mixed cellulose ester membrane filters.

@@

The generalized additive model (GAM) does better than the constant or even the linear model in accounting for autocorrelation in the data (Figure 2). This is because it accounts for non-linear patterns that the simpler models ignore [@trueman_seasonal_2022]. GAMs can also be thought of as linear models, but instead of regressing y on x, we regress y on *functions* of x (cubic functions, for example, as in Figure 3).

@@

$$(4)~~h(s, \phi) = \phi^s, s\geq0, \phi\geq0, \phi\leq1 $$

@@

Model code was written in Stan [@stan2021; @rstan], a C++ package for Bayesian inference using Markov chain Monte Carlo (MCMC) methods. We generated a code template using the R package `brms` [@brms; @brms_adv] and modified it to specify the continuous-time autocorrelation structure defined in equation $(4)$. We used a suite of common R packages [@tidyverse; @nada; @assertr; @janitor; @datatable], and bespoke functions used to fit the model and process the results are documented on GitHub [@bgamcar1]. The data and R code necessary to reproduce the analysis are available in a separate repository [@compare_cct].

```{r figure-4, fig.cap="**Figure 4. (a)** Seasonal component of the trend in dissolved lead. **(b)** Global longterm component of the trend in dissolved and particulate lead. **(c)** Group-level (i.e., pipe-specific) longterm trends in dissolved and particulate lead, representing the deviations of each time series from the global longterm trend. In panels **(a--b)**, shaded regions span the middle 95% of the posterior distribution, and 200 draws from that posterior are superimposed in blue. All plots are shown on the transformed scale and centered."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-4.png"))
```

The models included one-dimensional global smooths describing longterm variation in lead across all time series (Figure 4b). The model of dissolved lead also included a global smooth describing seasonal variation (Figure 4a). Both models included group-level smooths describing longterm variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed in a recent paper [@pedersen_hierarchical_2019]. The longterm smooths were thin-plate regression splines and the seasonal smooth was a cyclic cubic regression spline [@mgcv]. To speed up computation, the response variable was log-transformed and then scaled to have zero mean and unit variance.

@@

$$
(5)
\begin{align}
y_t | censored_t = 0 \sim T(\mu_t,\sigma,\nu) \\ 
y_t | censored_t = 1 \sim F(\mu_t,\sigma,\nu) \\
y_t | censored_t = 2 \sim G(\mu_t,\sigma,\nu) \\
\mu_t = \alpha_{pipe_i} + \sum_{j=1}^{n}f_j(t) + \phi^s r_{t-s} \\
\sigma \sim T(0, 2.5, 3) \\
\nu \sim Gamma(2,0.5) \\
\phi \sim N(0.5, 0.25) \\
\end{align}
$$

@@

$$
(6)
\begin{align}
\alpha_{pipe_i} \sim N(\alpha, \sigma_\alpha) \\
\alpha \sim T(0.1, 2.5, 3) \\
\sigma_{\alpha} \sim T(0, 2.5, 3) \\ 
\end{align}
$$
The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$, and $r_{t-s}$ is the GAM residual at the previous time step, $t-s$ ($r_{t-s} = y_{t-s} - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s)$. The $f_j(t)$ are smooth spline functions that comprise the GAM, and each of the $n$ smooths takes the following form:

$$
(7)
\begin{align}
f_j(t) = X_j\beta_j + Z_jb_j \\
\beta_j \sim T(0, 2.5, 3) \\
b_j \sim N(0, \sigma_b) \\
\sigma_b \sim N(0, 0.5) (seasonal~smooth) \\
\sigma_b \sim T(0, 2.5, 3) (all~other~smooths) \\ 
\end{align}
$$

where $Z_j$ and $X_j$ are matrices representing the penalized and unpenalized basis functions evaluated at time $t$; and $\beta$ and $b_j$ represent the corresponding spline coefficient vectors ($\beta$ is $1\times1$, corresponding to a single unpenalized basis function associated with the global longterm trend). The priors on $\sigma_b$ help to prevent overfitting by penalizing complexity in the fitted smooths [@mgcv]. And while the chosen priors did help with model convergence, they did not exert a strong influence on the posterior (Figure S2); the likelihood---representing `r nrow(model_in)` observations in each model---was overwhelmingly influential. A summary of the fitted models, including convergence statistics and effective sample sizes, is provided in Figure S3 and S4.

@@

We estimated the local slope of the global longterm smooths using a finite difference method. Specifically, we generated 4000 posterior predictions of the smooth trend along a regular sequence spanning the date range of the data. This was repeated after adding $\delta = 0.001$ (~$1/3$ of a day) to each value in the sequence. The difference between the grid of posterior predictions evaluated at $t$ and $t+\delta$, divided by $\delta$, approximates the instantaneous rate of change after accounting for seasonal variation, inter-pipe variation, and autocorrelation.

@@

# Results

@@

Time series of particulate and dissolved lead featured prominent negative trends in the first year of operation followed by an extended period of relative stability. The dissolved lead time series were seasonal, with maximum levels in late summer and minimum levels in early spring (Figure 4a). To estimate the time-varying mean representing each series, we modeled lead as the sum of a global longterm smooth, a global seasonal smooth (dissolved lead only), a group-level longterm smooth, a local intercept for each pipe, and a continuous-time autocorrelated error term (Figure 6).

@@

Effluent lead concentrations in pipe loop studies typically exhibit a pronounced negative trend after commissioning, followed by relative stability for the remainder of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant. This general pattern is represented in our model by the global longterm trend (Figure 4b), which excludes variation in effluent lead due to non-constant inputs (e.g., water quality and temperature). The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term. Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a). Both fractions exhibited positive and negative trends in subsequent years that are difficult to explain, but they may be a response to changes in water quality over the course of the study.

```{r figure-8, fig.cap="**Figure 8. (a) ** The global longterm smooth and its estimated local slope. A random sample of 200 smooths/slope estimates is superimposed, and both sets of lines are underlain by shaded regions covering the middle 95% of the respective posterior distributions. The median of the global trend is shaded red when the 95% credible interval around the local slope does not include zero. **(b--c)** Iron and water temperature, both measured in pipe rack influent, exhibited comparable seasonal patterns to lead, the latter described by the seasonal term in the GAM. Ticks at the top and bottom edges of the plot in **(b)** represent anomalously high and left-censored iron concentrations, respectively."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-8.png"))
```

@@

Over the course of the study, a majority of total lead was present in 0.45 µm filtrate (Figure 9). Orthophosphate treatment was associated with a larger particulate fraction, as is particularly evident in the first year of the study, where a comparison of 0, 1, and 2 mg P L^-1^ is possible. The particulate fraction was also greater in the effluent from pipes featuring a galvanic junction, and both patterns are consistent with previous work [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018].

>>>>

Pipe loop studies are used to evaluate corrosion control treatment, and updated regulatory guidance will ensure that they remain important for water quality management. But the data they generate are difficult to analyze: non-linear time-trends, non-detects, extreme values, and autocorrelation are common features that make popular methods, such as the *t*- or rank-sum tests, poor descriptive models. Here, we propose a model for pipe loop data that accommodates many of these difficult-to-model characteristics: a robust Bayesian generalized additive model with continuous-time autoregressive errors. Our approach facilitates corrosion control treatment comparisons without the need for imputing non-detects or special handling of outliers. It is well-suited to describing nonlinear trends without overfitting, and it accounts for the reduced information content in autocorrelated time series. We demonstrate it using a four-year pipe loop study, with multiple pipe configurations and orthophosphate dosing protocols, finding that an initially high dose of orthophosphate (2 mg P L^-1^) that is subsequently lowered (0.75 mg P L^-1^) can yield lower lead release than an intermediate dose (1 mg P L^-1^) in the long term.

@@

Pipe loop studies are used to model drinking water distribution systems, mainly for evaluating corrosion control treatment. The US EPA's Revised Lead and Copper Rule mandates them under certain conditions [@estessmargiassi_understanding_2020; @us2021national], and the data they generate will need to be analyzed. But pipe loop study data are inherently complex and there is a real risk of misinterpreting them. The following are just some of the challenges that arise (Figure 1):

@@

Left-censoring and autocorrelation can be particularly challenging to model. Left-censoring is often addressed by substituting a constant for values below the detection limit and sometimes by removing the censored values. Neither approach is advisable because both introduce artificial patterns that can distort results, sometimes seriously [@helsel_statistics_2012]. Autocorrelation overstates the precision of statistical estimates [@gelman_bayesian_2014]. As a consequence, it inflates the probability of falsely rejecting null hypotheses [@helsel1992statistical]. The effect of autocorrelation can be surprisingly large: 100 observations from a time series may provide no greater precision than 10 independent observations if the 100 are strongly correlated in time [@helsel1992statistical]. That is, *n* autocorrelated observations provide less information than *n* independent observations, and the difference can be large if autocorrelation is strong [@gelman_bayesian_2014]. Accounting for autocorrelation can improve predictive accuracy, but if it is ignored, analysts run the risk of making incorrect decisions based on the data. And while rank-based procedures such as the rank-sum test can accommodate left-censoring, most are not designed for autocorrelated data [@helsel1992statistical; @helsel_statistics_2012].

@@

1. A global---possibly nonlinear---trend describing variation in the parameter of interest over time across all pipes.
2. A set of group-level trends that describe the deviations from the global trend particular to each individual pipe.
3. A global seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature. (This component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it.)
4. An error term, possibly autocorrelated.

We demonstrate this approach using the data from a four-year pipe loop study that included multiple pipe configurations and orthophosphate dosing protocols. Our model estimates autocorrelation, despite an irregular sampling frequency; it also accommodates censored observations without imputation and is robust against extreme analyte concentrations due to particle release---meaning that no special treatment of outliers is necessary. An advantage of our Bayesian approach is the ability to move beyond point estimates and intervals: Bayesian models yield the full posterior distribution of all parameters, and all quantities derived from those parameters, including the predicted values. This allows for full propagation of uncertainty through all stages of post-processing and analysis [@beck_multi-scale_2022]. Moreover, the use of additive smoothing splines permits a description of time series using multiple nonlinear components, which can yield a more nuanced understanding of differences in lead concentrations over time.

@@

Lead, iron, aluminum, and manganese were measured by ICP-MS [@sm_3125] with method detection limits estimated at 0.2, 5, 5, and 2 µg L^-1^, respectively. An aliquot of each sample was filtered at 0.45 µm using a teflon filter and a vacuum filter apparatus. Temperature was measured using an infrared thermometer, total chlorine [@sm_4500_cl_g] and orthophosphate [@sm_4500_p_e] were measured using a spectrophotometer, and pH was measured using a benchtop meter. Water quality---including alkalinity [@sm_2320b], true colour [@sm_2120c], conductivity [@sm_2510b], fluoride [@sm_4500], total hardness [@sm_2340], and turbidity [@sm_2130b] ---is summarized in Table S4. Colour was quantified after filtration at 0.45 µm, using mixed cellulose ester membrane filters.

@@

The generalized additive model (GAM) does better than the constant or even the linear model in accounting for autocorrelation in the data (Figure 2). This is because it accounts for non-linear patterns that the simpler models ignore [@trueman_seasonal_2022; @murphy_generalized_2019]. GAMs can also be thought of as linear models, but instead of regressing y on x, we regress y on *functions* of x (cubic functions, for example, as in Figure 3).

@@

$$(4)~~h(s, \phi) = \phi^s, s\geq0, 0 \leq \phi \leq1 $$

@@

Model code was written in Stan [@stan2021; @rstan], a C++ package for Bayesian inference using Markov chain Monte Carlo (MCMC) methods. We generated a code template using the R package `brms` [@brms; @brms_adv] and modified it to specify the continuous-time autocorrelation structure defined in equation $(4)$. We used a suite of R packages [@tidyverse; @nada; @assertr; @janitor; @datatable; @patchwork; @posterior; @ggdist; @wesanderson; @ggh4x; @broom; @ggtext], and bespoke functions used to fit the model and process the results are documented on GitHub [@bgamcar1]. The data and R code necessary to reproduce the analysis are available in a separate repository [@compare_cct].

```{r figure-4, fig.cap="**Figure 4. (a)** Seasonal component of the trend in dissolved lead. **(b)** Global long-term component of the trend in dissolved and particulate lead. **(c)** Group-level (i.e., pipe-specific) long-term trends in dissolved and particulate lead, representing the deviations of each time series from the global long-term trend. In panels **(a--b)**, shaded regions span the middle 95% of the posterior distribution, and 200 draws from that posterior are superimposed in blue. All plots are shown on the transformed scale and centered."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-4.png"))
```

The models included one-dimensional global smooths describing long-term variation in lead across all time series (Figure 4b). The model of dissolved lead also included a global smooth describing seasonal variation (Figure 4a). Both models included group-level smooths describing long-term variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed in a recent paper [@pedersen_hierarchical_2019]. The long-term smooths were thin-plate regression splines and the seasonal smooth was a cyclic cubic regression spline [@mgcv]. To speed up computation, the response variable was log-transformed and then scaled to have zero mean and unit variance.

@@

$$
(5)~~
\begin{align}
y_t | censored_t = 0 \sim T(\mu_t,\sigma,\nu) \\ 
y_t | censored_t = 1 \sim F(\mu_t,\sigma,\nu) \\
y_t | censored_t = 2 \sim G(\mu_t,\sigma,\nu) \\
\mu_t = \alpha_{pipe_i} + \sum_{j=1}^{n}f_j(t) + \phi^s r_{t-s} \\
\sigma \sim T(0, 2.5, 3) \\
\nu \sim Gamma(2,0.5) \\
\phi \sim N(0.5, 0.25) \\
\end{align}
$$

@@

$$
(6)~~
\begin{align}
\alpha_{pipe_i} \sim N(\alpha, \sigma_\alpha) \\
\alpha \sim T(0.1, 2.5, 3) \\
\sigma_{\alpha} \sim T(0, 2.5, 3) \\ 
\end{align}
$$
The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$, and $r_{t-s}$ is the GAM residual at the previous time step, $t-s$; $r_{t-s} = y_{t-s} - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s)$. The $f_j(t)$ are smooth spline functions that comprise the GAM, and each of the $n$ smooths takes the following form:

$$
(7)~~
\begin{align}
f_j(t) = X_j\beta_j + Z_jb_j \\
\beta_j \sim T(0, 2.5, 3) \\
b_j \sim N(0, \sigma_b) \\
\sigma_b \sim N(0, 0.5) (seasonal~smooth) \\
\sigma_b \sim T(0, 2.5, 3) (all~other~smooths) \\ 
\end{align}
$$

where $Z_j$ and $X_j$ are matrices representing the penalized and unpenalized basis functions evaluated at time $t$; and $\beta$ and $b_j$ represent the corresponding spline coefficient vectors ($\beta$ is $1\times1$, corresponding to a single unpenalized basis function associated with the global long-term trend). The priors on $\sigma_b$ help to prevent overfitting by penalizing complexity in the fitted smooths [@mgcv]. And while the chosen priors did help with model convergence, they did not exert a strong influence on the posterior (Figure S2); the likelihood---representing `r nrow(model_in)` observations in each model---was overwhelmingly influential. A summary of the fitted models, including convergence statistics and effective sample sizes, is provided in Figure S3 and S4.

@@

We estimated the local slope of the global long-term smooths using a finite difference method. Specifically, we generated 4000 posterior predictions of the smooth trend along a regular sequence spanning the date range of the data. This was repeated after adding $\delta = 0.001$ (~$1/3$ of a day) to each value in the sequence. The difference between the grid of posterior predictions evaluated at $t$ and $t+\delta$, divided by $\delta$, approximates the instantaneous rate of change after accounting for seasonal variation, inter-pipe variation, and autocorrelation.

@@

# Results and discussion

@@

Time series of particulate and dissolved lead featured prominent negative trends in the first year of operation followed by an extended period of relative stability. The dissolved lead time series were seasonal, with maximum levels in late summer and minimum levels in early spring (Figure 4a). To estimate the time-varying mean representing each series, we modeled lead as the sum of a global long-term smooth, a global seasonal smooth (dissolved lead only), a group-level long-term smooth, a local intercept for each pipe, and a continuous-time autocorrelated error term (Figure 6).

@@

Effluent lead concentrations in pipe loop studies typically exhibit a pronounced negative trend after commissioning, followed by relative stability for the remainder of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant. This general pattern is represented in our model by the global long-term trend (Figure 4b), which excludes variation in effluent lead due to non-constant inputs (e.g., water quality and temperature). The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term. Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a). Both fractions exhibited positive and negative trends in subsequent years that are difficult to explain, but they may be a response to changes in water quality over the course of the study.

```{r figure-8, fig.cap="**Figure 8. (a) ** The global long-term smooth and its estimated local slope. A random sample of 200 smooths/slope estimates is superimposed, and both sets of lines are underlain by shaded regions covering the middle 95% of the respective posterior distributions. The median of the global trend is shaded red when the 95% credible interval around the local slope does not include zero. **(b--c)** Iron and water temperature, both measured in pipe rack influent, exhibited comparable seasonal patterns to lead, the latter described by the seasonal term in the GAM. Ticks at the top and bottom edges of the plot in **(b)** represent anomalously high and left-censored iron concentrations, respectively."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-8.png"))
```

@@

Over the course of the study, a majority of total lead was present in 0.45 µm filtrate (Figure 9). Orthophosphate treatment was associated with a larger particulate fraction, as is particularly evident in the first year of the study, where a comparison of 0, 1, and 2 mg P L^-1^ is possible. The particulate fraction was also greater in the effluent from pipes featuring a galvanic junction, and both patterns are consistent with previous work [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017].
